---
title: "ggplot2 examples (Bio-data Club)"
author: "Ram Thapa"
date:  "2020-08-21"
output: 
  html_document:
    df_print: paged
    code_folding: show
    toc: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 5, fig.align = "center", warning = FALSE)
```

Data visualization is important in most phases of data analysis workflow and being able to create visualizations of data is a key step in being able to effectively communicate information and findings to others. 

{ggplot2} is a plotting framework that is (relatively) easy to use, powerful and hassle-free. It is based on a set of principles "Grammar of graphic" and is designed to work in a layered fashion, starting with a layer showing the data then adding layers of graphical objects, annotations and statistical summaries.

**Packages and data**

```{r packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggridges)
library(patchwork)

head(penguins)
```

## Visualizing associations among two or more variables

### Scatterplot

Relationship between bill length and flipper length

```{r scatterplot}
# set theme to theme_bw()
theme_set(theme_bw(base_size = 15))

ggplot(data = penguins, aes(x = bill_length_mm, y = flipper_length_mm)) +
  labs(x = "Bill length (mm)", y = "Flipper length (mm)") +
  geom_point(color = "gray30", size = 2.5)
```

Mapping `color` to `sex` variable from the data

```{r scatterplot2, fig.width = 7}
scatter_sex <- ggplot(data = penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = sex)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(x = "Bill length (mm)", y = "Flipper length (mm)") +
  scale_color_brewer(NULL, palette = "Set1", na.translate = FALSE) 

scatter_sex
```

Avoid overplotting using jitter plot

```{r jitter,  fig.width = 7}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm)) +
  geom_point() +
  geom_jitter(aes(color = sex), width = 0.2, size = 2, alpha = 0.7) +
  labs(x = NULL, y = "Flipper length (mm)") +
  scale_color_brewer(NULL, palette = "Set1", na.translate = FALSE) 
```

Note: `geom_jitter` is a shortcut for `geom_point(position = "jitter")`

### Line plot

Line plot of mean flipper length at each study year by species

```{r line, fig.width = 7}
line_species <- penguins %>% 
  group_by(species, year) %>% 
  summarise(mean = mean(flipper_length_mm, na.rm = TRUE)) %>% 
  ggplot(aes(x = as.character(year), y = mean, group = species, color = species)) +
  geom_point(size = 4) +
  geom_line(show.legend = FALSE) +
  scale_x_discrete("Study year", expand = c(0.1, 0.1)) +
  scale_y_continuous("Mean flipper length (mm)") +
  scale_color_brewer("Species", palette = "Dark2") 

line_species
```

### 2d density plot

A 2d density plot is useful to study the relationship between two numeric variables if there are huge number of points.

```{r density2d, fig.width = 7}
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_density_2d(show.legend = FALSE) +
  geom_point() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)") +
  scale_color_brewer("Species", palette = "Dark2")
```

## Visualizing amounts

We might be interested in visualizing amounts i.e. numerical values for some set of categories

### Bar plot

```{r barplot}
ggplot(data = penguins, aes(x = species)) +
  geom_bar(fill = "#4db39b")
```

You could also calculate counts in advance and plot it but you need to use `stat = "identity"`

```{r count}
# Count per species
species_count <- penguins %>% 
  count(species)

# use stat = "identity"
ggplot(data = species_count, aes(x = species, y = n)) +
  geom_bar(stat = "identity", fill = "#4db39b")
```

Or, just use `geom_col()`

```{r count2}
ggplot(data = species_count, aes(x = species, y = n)) +
  geom_col(fill = "#4db39b")
```

Sometimes x-axis labels identifying each bar take up a lot of horizontal space

```{r}
ggplot(species_count, aes(x = fct_reorder(species, n), y = n)) +
  geom_bar(stat = "identity", fill = "#4db39b") +
  labs(x = "Species", y = NULL) +
  coord_flip()
```

Grouped bar plot

```{r grouped_bar, fig.width = 7}
groupedbar_species <- ggplot(data = penguins, aes(x = year, fill = species)) +
  geom_bar(position = "dodge") +
  labs(x = NULL, y = "Number of penguins") +
  scale_fill_viridis_d("Species")

groupedbar_species
```

Stacked bar plot

```{r stacked_bar, fig.width = 6}
ggplot(data = penguins, aes(x = year, fill = species)) +
  geom_bar(position = "stack") +
  labs(x = NULL, y = "Number of penguins") +
  scale_fill_viridis_d("Species")
```

Percent stacked bar plot

```{r percent_stacked_bar, fig.width = 6}
stackedbar_species <- ggplot(data = penguins, aes(x = year, fill = species)) +
  geom_bar(position = "fill") +
  labs(x = NULL, y = NULL) +
  scale_fill_viridis_d("Species") +
  scale_y_continuous(labels = scales::percent_format())

stackedbar_species
```

## Visualizing distributions

To understand how a particular variable is distributed in a dataset

### Boxplot

```{r boxplot}
ggplot(data = penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = NULL, y = "Body mass (gm)") +
  scale_fill_brewer(palette = "Dark2")
```

### Violin plot

Similar to box plots except they also show the kernel probability density of the data at different values
 
```{r violin, fig.width = 8}
ggplot(data = penguins, aes(x = factor(year), y = body_mass_g, fill = species)) +
  geom_violin(trim = FALSE) +
  labs(x = NULL, y = "Body mass (gm)") +
  scale_fill_brewer("Species", palette = "Dark2")
```


### histogram and density plot

Histogram count plot for single distribution

```{r histogram}
hist_flipper <- ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(fill = "#4db39b", color = "#e9ecef") +
  labs(x = "Flipper length (mm)")

hist_flipper
```

Visualizing two distributions using pyramid plot

```{r pyramid, fig.width = 7}
ggplot(penguins, aes(x = flipper_length_mm, fill = sex)) +
  geom_histogram(
    aes(y = after_stat(ifelse(fill == unique(fill)[1], 1, -1) * count)),
    color = "gray80") +
  scale_x_continuous("Flipper length (mm)", expand = c(0, 0)) +
  scale_y_continuous("Count", breaks = seq(-20, 20, 5), labels = abs(seq(-20, 20, 5))) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  annotate("text", x = 230, y = c(-12, 10), label = c("Male", "Female"), size = 6) +
  coord_flip() +
  theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank())
```


Finally, when we want to visualize exactly two distributions, we can also make two separate histograms, rotate them by 90 degrees, and have the bars in one histogram point into the opposite direction of the other. This trick is commonly employed when visualizing age distributions, and the resulting plot is usually called an age pyramid (Figure 7.10).

Density plot for visualizing multiple distributions at the same time

```{r density, fig.width = 7}
ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.8, color = "transparent") +
  labs(x = "Flipper length (mm)") +
  scale_fill_viridis_d("Species") 
```

Use `after_stat()` to modify mapping from computed values by `stats`

```{r scaled_density, fig.width = 7}
ggplot(penguins, aes(x = flipper_length_mm, fill = species)) +
  geom_density(aes(y = after_stat(scaled))) +
  labs(x = "Flipper length (mm)", title = "Density scaled to 1") +
  scale_fill_viridis_d("Species")
```

### Ridge plot

Visualizing multiple distributions along the horizontal axis by staggering the distribution plots in the vertical direction

You need `{ggridges}` package for this plot

```{r ridge, fig.width = 7}
ridge_body_mass <- ggplot(
  penguins, aes(x = body_mass_g, y = fct_rev(factor(year)))) +
  geom_density_ridges(
    scale = 1.5, fill = "#39558CFF", color = "white") +
  scale_x_continuous("Body mass (gm)", expand = c(0, 0)) +
  scale_y_discrete(NULL, expand = c(0, 0.2, 0, 1.5)) +
  theme_minimal(base_size = 15)

ridge_body_mass
```

Comparing two distribution over time

```{r ridge2, fig.width = 7}
penguins %>% 
  drop_na(sex) %>% 
  ggplot(aes(x = body_mass_g, y = fct_rev(factor(year)), fill = sex)) +
  geom_density_ridges(
    scale = 1.5, rel_min_height = 0.01, color = "white", alpha = 0.8) +
  scale_x_continuous("Body mass (gm)", expand = c(0, 0)) +
  scale_y_discrete(NULL, expand = c(0, 0.2, 0, 1.7)) +
  scale_fill_brewer(NULL, palette = "Set1") +
  theme_minimal(base_size = 15)
```

## Combining plots

`{patchwork}` R package has made combining multiple plots ridiculuous simple

### Horizontal layout

```{r, fig.width = 8}
# combine two plots horizontally
scatter_sex + line_species
```

or you can use "|" too

### Vertical layout

```{r, message = FALSE, fig.height = 6}
# stack two plots
ridge_body_mass / hist_flipper
```

### Nested layout

```{r, message = FALSE, fig.height = 6, fig.width = 8}
patch <- line_species / (scatter_sex | hist_flipper)

patch
```

Be careful with the operator precedence rule (e.g. / is evaluated before |)

`{patchwork}` provides `wrap_plots()` for a more functional approach to assembly

### Annotating plots

```{r, message = FALSE, fig.height = 6, fig.width = 8}
(ridge_body_mass / (scatter_sex | hist_flipper)) +
  plot_annotation(
    title = "Combining three plots I made",
    subtitle = "Hopefully they look fine",
    caption = "Data source: {palmerpenguins} package",
    tag_levels = "A", tag_suffix = ".")
```   

### Modifying patches

Use `&` to add elements to all subplots in the patchwork

```{r, message = FALSE, fig.height = 6, fig.width = 8}
patch & theme_gray()
```

and `*` to add the element to all the subplots in the current nesting level

```{r, message = FALSE, fig.height = 6, fig.width = 8}
patch * theme_gray()
```

### Controlling guides

If multiple plots have same guide, using `guides = "collect"` in `plot_layout()` drops the duplicate guide

```{r, message = FALSE, fig.height = 6, fig.width = 8}
(groupedbar_species | (stackedbar_species / hist_flipper)) +
  plot_layout(guides = "collect")
```
